Global code motion of congruent instructions

### 2017-04-18 paper.tex
grammar

There have been a lot of work both in industry and academia to hoist and sink
code out of branches, and in general global scheduling
\cite{click1995global}. Some relate code-hoisting to code-size optimization
\cite{rosen1988global} and many \cite{barany2013, shobaki2013} use global
scheduling to improve performance. Most of the recent work on global scheduling
are done using integer linear programming (ILP), which results in prohibitively
high compile time. To the best of our knowledge we have not found any reference
which explored \gcm{} of congruent instructions in as much detail as we
have done. A part of our implementation (aggressive code hoisting) is already
merged in \LLVM{} trunk, however, a general implementation of \gcm{} of
congruent (non redundant) expressions is still missing from \GCC{} and \LLVM{}
trunk. The main contributions of this paper are:
\begin{itemize}
\item a new optimistic algorithm to move congruent instructiosn out of branches;
\item a cost model to reduce interference and hence, reduce spills;
\item a technique to maximize hoisting in an optimistic approach by partitioning
  the list of potential candidates sorted by their DFS visit number;
\item experimental evaluation of our implementation in \LLVM{} which combines
  \SSA{} based liveness analysis, and ranking of expressions to move very busy
  expressions in order to reduce register spills and improve performance in some
  cases.
\end{itemize}

@@

We compute merge sets of the control flow graph which remains the same through
the \gcm{} transformation. For \gcm{} we only want to know if a use
operand is a kill (to compute changes in register pressure.) For that we only
need to know whether the use is also required later in the execution path. A
variable is live out of a basic block B if it is used in the merge set of B. We
compute the live-out relation on-demand when profitability of hoistable/sinkable
candidates is to be evaluated. A simplified version of
\emph{isLiveOutUsingMergeSet} is presented here:

@@

The \GCM{} pass can be broadly divided into the following steps that we will
describe in the rest of this section:
\begin{itemize}
\item find candidates (congruent instructions) suitable for \gcm{} (Section~\ref{subsec:finding-candidates}),
\item compute a point in the program where it is both legal
  (Section~\ref{subsec:legality}) and profitable
  (Section~\ref{subsec:cost-models}) to move the code,
\item move the code to hoist point or sink point
  (Section~\ref{subsec:optimistic}), and
\item update data structures to continue iterative \gcm{}.
\end{itemize}

@@

\subsection{Time complexity of algorithm}
The complexity of code hoisting is linear in number of instructions that are
candidates for \gcm{}, matching the complexity of \PRE{} on \SSA{} form.
The analyses computed for this pass are Global Value Numbering, Computation of
DJ-graph and merge-sets, Marking Barriers, all are linear in number of
instructions in a function. Liveness analysis is expensive and only performed
on-demand for hoistable candidates so it does not affect the overall compilation
time by much. Other analyses like Alias Analysis, \MemorySSA{} and Dominator Tree
are already available in the \LLVM{} pass pipeline. Although we recompute \GVN{}, and
Barriers for each iteration of the \gcm{}, we have not seen signinficant
increase in the compilation times. We have also provided appropriate compiler
flags to expedite \gcm{} by bailing out with fewer iterations, or skip the
liveness based profitability analysis to aggressively move the code as long as
they are legal (Section~\ref{sec:experimental-results}).

@@

\section{Experimental Evaluation}
\label{sec:experimental-results}
% Experimental results: Register spills, compile time, run time. compile-time comparison w.r.t. gcc
We ran \SPEC{} with the patch. All the experiments were conducted on an
\xlinux{} machine and at -Ofast optimization level. Each sub benchmark was
run three times and the best result was taken. We collected execution time,
code-size which is listed in Table~\ref{tab:code-size}, other compiler stats are
listed in Table~\ref{tab:spills}, Table~\ref{tab:stats}, and
Table~\ref{tab:code-motion-metric}.  We also measured compile time
Table~\ref{tab:compile-time} to see the impact of \gcm{}.

@@

Overall there was only a minor change in performance regressed with a few ups
(429.mcf, 458.sjeng etc.) and a few regresssions (447.dealII, 453.povray
etc.). The \gcm{} pass was run twice in the pass pipeline (at -Ofast),
first after EarlyCSE pass, and then after the inliner. Because EarlyCSE removes
local redundancies, \GCM{} would have to analyze less redundant instructions, also
as inliner creates more redundancies, \GCM{} would have more opportunities to move
congruent instructions out of branches.

@@

The compile time metrics of \GCM{} are listed in
Table~\ref{tab:code-motion-metric}. The table lists the number of scalars,
loads, stores and calls hoisted as well as removed. For each category, the
number of instructions removed is greater or equal to the number of instructions
hoisted because each hoisting is performed only when at least one identical
computation is found. Loads are hoisted the most followed by scalars, stores and
calls in decreasing order.  This was the common trend in all our
experiments. One reason why loads are hoisted the most is the early execution of
this pass (before mem2reg) in the \LLVM{} pass pipeline. Passes like mem2reg and
instcombine might actually remove those loads and the number of hoisted loads
may change should the \GCM{} pass be scheduled later. We can see a few sunk
instructions even if \LLVM{} has a code sinking and hoisting transforms in the
SimplifyCFG pass which runs before \GCM{}: this is because sinking and hoisting
in SimplifyCFG have some rather severe limitations to make the analysis and code
transform very fast in order to be able to run the pass several times.  Another
reason is that instructions for which dependencies are not directly available in
the successor (Section~\ref{subsec:legality-sink}) is not sunk for now. Sinking
those instructions would require a look ahead on the ability to sink the
dependencies. We plan to implement this in near future.

@@

The compilation time actually reduced for most of the benchmarks as a result of code motion.
This is because as \gcm{} removes (congruent) instructions, the number of instructions
to be processed by later compilation passes also reduces. The compilation time listed here
is the total instruction count as outputted from valgrind (--tool=cachegrind).

@@

Some useful statc metric are presented in Table~\ref{tab:stats}, which shows how \gcm{}
impacted important compiler optimizations like inlining and LICM. LICM removes loop invariant
code out of the loop. It may hoist the code to the loop's preheader or sink the code to loop's
post-dominator \cite{steven1997advanced}. As we can see number of instructions hoisted increased
significantly in several benchmarks, although number of instructions sunk did not change by much.
Number of inlined functiosn also improved for most benchmarks. In general when more functions
were inlined, more functions were deleted as in 400.perlbench, 445.gobmk.

@@

While benchmarking \SPEC{} we see both increase as well as decrease in
the codesizes of the final binaries. Since the pass runs early, it affects many
optimizations which rely on the number of instructions, lenth of the use-def
chain, and other metrics. For instance, the inliner is impacted by a decrease in
the number of instructions in the caller and callee, as its heuristics estimate
the size of functions to be inlined.

@@

\section{Conclusion and Future Work}
\label{sec:future-work}
We have presented the \gcm{} of congruent computations in SSA form. We saw that
it improves inlining and LICM opportunities, reduces register spills, it also
improves performance in some cases. To preserve performance and not hoist too
much we have presented some cost models to lower register pressure. A part of
this implementation is already merged in LLVM trunk as GVNHoist.cpp, and we
posted the \GCM{} implementation for review.

There are a few improvements that we would like to make. First would be to
integrate this with \LLVM{'}s new GVN-PRE implementation such that more
congruent instructions can be scheduled. This would also expose more
redundancies to the GVN-PRE.  In general it is a good idea to start with lower
ranked expression first such that maximum hoisting can happen in one iteration,
however, current implementation does not rank the expressions and iteratively
finds a fixed point when no more candidates are available. Even this
implementation converges quickly and no compile time regressions have been
observed because of code hoisting pass. This is not the most optimal approach
and results in multiple data structures to be recomputed. This can be improved
by ranking the computations \cite{rosen1988global}. Also, since \GCM{} runs very
early in the pass pipeline, it will be good to evaluate the codesize/performance
impact when it is run in sync with \GVN{}-\PRE{} just like \GCC{} does.

With the implementation of \GCM{} in \LLVM{}, the passes which rely on the
code-size/instruction-count to make optimization decisions need to be
revisited. The first candidate would be the inliner. We have seen different
inlining decisions in Table~\ref{tab:code-size}, before and after code-hoisting
was enabled.  Since inliner has several magic numbers tuned for the previous
pass layout, it would need some improvement.

>>>>

There has been a lot of work both in industry and academia to hoist and sink
code out of branches, and in general global scheduling
\cite{click1995global}. Some relate code-hoisting to code-size optimization
\cite{rosen1988global} and many \cite{barany2013, shobaki2013} use global
scheduling to improve performance. Most of the recent work on global scheduling
are done using integer linear programming (ILP), which results in prohibitively
high compile time. To the best of our knowledge we have not found any reference
which explored \gcm{} of congruent instructions in as much detail us. A part of
our implementation (aggressive code hoisting) is already merged
in \LLVM{} trunk, however, a general implementation of \gcm{} of congruent (non
redundant) expressions is still missing from \GCC{} and \LLVM{} trunk. The main
contributions of this paper are:
\begin{itemize}
\item a new optimistic algorithm to move congruent instructions out of branches;
\item a cost model to reduce interference and hence, reduce spills;
\item a technique to maximize hoisting in an optimistic approach by partitioning
  the list of potential candidates sorted by their DFS visit number;
\item experimental evaluation of our implementation in \LLVM{} which combines
  \SSA{} based liveness analysis, and ranking of expressions to move very busy
  expressions in order to reduce register spills and improve performance in some
  cases.
\end{itemize}

@@

We compute merge sets, of the control flow graph, which remain same throughout
the \gcm{} transformation. For \gcm{} we only want to know if a use
operand is a kill (to compute changes in register pressure.) For that we only
need to know whether the use is also required later in the execution path. A
variable is live out of a basic block B if it is used in the merge set of B. We
compute the live-out relation on-demand when profitability of hoistable/sinkable
candidates is to be evaluated. A simplified version of
\emph{isLiveOutUsingMergeSet} is presented here:

@@

The \GCM{} pass can be broadly divided into the following steps that we will
describe in the rest of this section:
\begin{itemize}
\item find candidates (congruent instructions) suitable for \gcm{} (Section~\ref{subsec:finding-candidates}),
\item compute a point in the program where it is both legal
  (Section~\ref{subsec:legality}) and profitable
  (Section~\ref{subsec:cost-models}) to move the code,
\item move the code to hoist point or sink point
  (Section~\ref{subsec:optimistic}), and
\item update data structures to continue iterative \gcm{} (Section~\ref{subsec:optimistic}).
\end{itemize}

@@

\subsection{Time complexity of algorithm}
The complexity of code hoisting is linear in number of instructions that are
candidates for \gcm{}, matching the complexity of \PRE{} on \SSA{} form.
The analyses computed for this pass are Global Value Numbering, Computation of
DJ-graph and merge-sets, Marking Barriers, all are linear in number of
instructions in a function. Liveness analysis is expensive and only performed
on-demand for hoistable candidates so it does not affect the overall compilation
time by much. Other analyses like Alias Analysis, \MemorySSA{} and Dominator Tree
are already available in the \LLVM{} pass pipeline. Although we recompute \GVN{}, and
Barriers for each iteration of the \gcm{}, we have not seen signinficant
increase in the compilation times (see Section~\ref{tab:compile-time}). We have also
provided appropriate compiler
flags to expedite \gcm{} by bailing out with fewer iterations, or skip the
liveness based profitability analysis to aggressively move the code as long as
they are legal.

@@

\section{Experimental Evaluation}
\label{sec:experimental-results}
% Experimental results: Register spills, compile time, run time. compile-time comparison w.r.t. gcc
For evaluation of \gcm{}, we built \SPEC{} with the patch. All the experiments
were conducted on an \xlinux{} machine and at -Ofast optimization level. Each
sub benchmark was run three times and the best result was taken. We collected
execution time and code-size which is listed in Table~\ref{tab:code-size}, other
compiler stats are listed in Table~\ref{tab:spills}, Table~\ref{tab:stats}, and
Table~\ref{tab:code-motion-metric}.  We also measured compile time
Table~\ref{tab:compile-time} to see the impact of \gcm{}.

@@

Overall there was only a minor change in the performance numbers with a few improvements 
(429.mcf, 458.sjeng etc.) and a few regresssions (447.dealII, 453.povray
etc.). The \gcm{} pass was run twice in the pass pipeline (at -Ofast),
first after EarlyCSE pass, and then after the inliner. Because EarlyCSE removes
local redundancies, \GCM{} would have to analyze less redundant instructions, also
as inliner creates more redundancies, \GCM{} would have more opportunities to move
congruent instructions out of branches.

@@

The compile time metrics of \GCM{} are listed in
Table~\ref{tab:code-motion-metric}. The table lists the number of scalars,
loads, stores and calls hoisted as well as removed. For each category, the
number of instructions removed is greater or equal to the number of instructions
hoisted because each code-motion is performed only when at least one identical
computation is found. Loads are hoisted the most followed by scalars, stores and
calls in decreasing order.  This was the common trend in all our
experiments. One reason why loads are hoisted the most is the early execution of
this pass (before mem2reg pass which scalarizes some memory references) in the \LLVM{} pass pipeline. Passes like mem2reg and
instcombine might actually remove those loads and the number of hoisted loads
may change should the \GCM{} pass be scheduled later. We can see a few sunk
instructions even if \LLVM{} has a code sinking and hoisting transforms in the
SimplifyCFG pass which runs before \GCM{}: this is because sinking and hoisting
in SimplifyCFG have some rather severe limitations to make the analysis and code
transform very fast in order to be able to run the pass several times.  Another
reason is that instructions for which dependencies are not directly available in
the successor (Section~\ref{subsec:legality-sink}) is not sunk for now. Sinking
those instructions would require a look ahead on the ability to sink the
dependencies. We plan to implement this in near future.

@@

The compilation time actually reduced for most of the benchmarks as a result of code motion.
This is because, as \gcm{} removes (congruent) instructions, the number of instructions
to be processed by later compilation passes also reduces. The compilation time listed here
is the total instruction count as outputted from valgrind (--tool=cachegrind) \cite{valgrind}.

@@

Some useful static metrics are presented in Table~\ref{tab:stats}, which shows how \gcm{}
impacted important compiler optimizations like inlining and LICM. LICM removes loop invariant
code out of the loop. It may hoist the code to the loop's preheader or sink the code to loop's
post-dominator \cite{steven1997advanced}. As we can see number of instructions hoisted increased
significantly in several benchmarks, although number of instructions sunk did not change by much.
Number of inlined functions also improved for most benchmarks. In general when more functions
were inlined, more functions were deleted as in 400.perlbench, 445.gobmk.

@@

While benchmarking \SPEC{} we see both increase as well as decrease in
the codesize of the final binaries. Since the pass runs early, it affects many
optimizations which rely on the number of instructions, length of the use-def
chain, and other metrics. For instance, the inliner is impacted by a decrease in
the number of instructions in the caller and callee, as its heuristics estimate
the size of functions to be inlined.

@@

\section{Conclusion and Future Work}
\label{sec:future-work}

We have presented the \gcm{} of congruent computations in SSA form. We saw that
it improves inlining and LICM opportunities, reduces register spills, it also
improves performance in some cases. To preserve performance and not hoist/sink
too much, we have implemented a register pressure aware cost model as described
in Section~\ref{subsec:cost-models}. A part of this implementation is already
merged in LLVM trunk as GVNHoist.cpp and we posted the \GCM{} implementation for
review

There are a few improvements that we would like to make. First would be to
integrate this with \LLVM{'}s new GVN-PRE implementation such that more
congruent instructions can be scheduled. This would also expose more
redundancies to the GVN-PRE.  In general it is a good idea to start with lower
ranked expressions first such that maximum hoisting can happen in one iteration,
however, current implementation does not rank the expressions and iteratively
finds a fixed point when no more candidates are available. Even this
implementation converges quickly and no compile time regression have been
observed. This is not the most optimal approach in terms of compile time complexity
and results in multiple data structures to be recomputed. This can be improved
by ranking the computations \cite{rosen1988global}. Also, since \GCM{} runs very
early in the pass pipeline, it will be good to evaluate the codesize/performance
impact when it is run in sync with \GVN{}-\PRE{} just like \GCC{} does.

With the implementation of \GCM{} in \LLVM{}, the passes which rely on the
code-size or instruction-counts to make optimization decisions need to be
revisited. The first candidate would be the inliner. We have seen difference
in inlining decisions (Table~\ref{tab:code-size}), when \gcm{}
was enabled.  Since inliner has several cost-models tuned for the previous
pass layout, that would need some improvement.
